Non-parametric approaches to optimal policy are more robust

### 2013-04-02 inst/doc/nonparametric-bayes.Rmd
writing writing

```{r knit-settings, cache=FALSE}
options(xtable.print.comment=FALSE)
options(xtable.type = 'latex', table.placement="H")
opts_chunk$set(warning=FALSE, message=FALSE, comment=NA, tidy=FALSE,
 echo=FALSE)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(dev = 'Cairo_pdf', fig.width=5.5, fig.height=4,
 cache.path = 'cache-pdf/', cache=TRUE)
```

```{r libraries, include=FALSE}
library(pdgControl)
library(nonparametricbayes)
library(reshape2)
library(ggplot2)
library(data.table)
library(tgp)
library(MCMCpack)
library(plyr)
library(knitcitations)
```
```{r plotting-options, echo=FALSE, include=FALSE}
theme_set(theme_bw(base_size=10))
theme_update(panel.background = element_rect(fill = "transparent",colour = NA),
             plot.background = element_rect(fill = "transparent",colour = NA))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

@@

@@

@@

@@

@@

### The optimal control problem

@@

Concerns over the potential for tipping points in ecological dynamics
[@Scheffer2001] highlight the dangers of uncertainty in ecological
management and pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  To compare the performance of nonparametric
and parametric approaches in an example that is easy to conceptualize,
we will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".

@@

## Sample training data

@@

<span style="color:#007FFF;">***Should we include any emprical examples?***</span>
Using data simulated from a specified model rather than empirical data
permits the comparison against the true underlying dynamics, setting 
a bar for the optimal performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

### Discussion of maximum likelihood estimated models

```{r obs, dependson="sdp-pars"}
  obs <- sim_obs(Xo, z_g, f, p, Tobs=Tobs, nz=15, 
                 harvest = sort(rep(seq(0, .5, length=7), 5)), seed = seed_i)
```

(MLE models will assume the noise is log-normal, which it is in the
simulation).

Which estimates a Ricker model with $r =$ `r alt$p[1]`, $K =$ 
`r alt$p[2]`, and the Allen Allele model with $r =$ `r est$p[1]`, $K =$
`r est$p[2]` and $C =$ `r est$p[3]`.

## Figure 1:

_Shows the inferred Gaussian Process compared to the true and parametric
models.  Refer to the appendix for details on the GP posteriors, etc._

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). "}
tgp_dat <- 
    data.frame(  x = gp$XX[[1]], 
                 y = gp$ZZ.km, 
                 ymin = gp$ZZ.km - 2 * sqrt(gp$ZZ.ks2), 
                 ymax = gp$ZZ.km + 2 * sqrt(gp$ZZ.ks2),
                 ymin2 = gp$ZZ.mean - 2 * sqrt(gp$ZZ.vark), 
                 ymax2 = gp$ZZ.mean + 2 * sqrt(gp$ZZ.vark))
  true <- sapply(x_grid, f, 0, p)
  alt_mean <- sapply(x_grid, alt$f, 0, alt$p)
  est_mean <- sapply(x_grid, est$f, 0, est$p)
  models <- data.frame(x=x_grid, GP=tgp_dat$y, 
                       Parametric=est_mean, 
                       True=true, 
                       Structural=alt_mean)
  models <- melt(models, id="x")
  names(models) <- c("x", "method", "value")

ggplot(tgp_dat) + 
  geom_ribbon(aes(x,y,ymin=ymin,ymax=ymax), fill="gray80") +
  geom_ribbon(aes(x,y,ymin=ymin2,ymax=ymax2), fill="gray60") +
  geom_line(data=models, aes(x, value, col=method), alpha=0.8, lwd=1) + 
  geom_point(data=obs, aes(x,y)) + 
  xlab(expression(X[t])) + ylab(expression(X[t+1])) +
  scale_colour_manual(values=cbPalette) 
```

@@

parameter       interpretation
---------       -------------- 
$\sigma^2$      The process noise (from the kernel)
$\tau^2$        Variance around the mean
$\beta$_0       The mean is given by a linear model of slope $\beta$ 
$d$             The length-scale of the covariance function
$n$             The observation error

>>>>

```{r libraries, include=FALSE}
## Dargh these are ignored if not placed here!?
options(xtable.type = 'latex', table.placement="H", xtable.print.comment=FALSE)
opts_chunk$set(cache.path = 'cache-pdf/', cache=TRUE)
opts_chunk$set(dev = 'Cairo_pdf', fig.width=5.5, fig.height=4)

library(pdgControl)
library(nonparametricbayes)
library(reshape2)
library(ggplot2)
library(data.table)
library(tgp)
library(MCMCpack)
library(plyr)
library(knitcitations)
```

```{r plotting-options, echo=FALSE, include=FALSE}
# This stuff should be handled by knit script(?) (load ggplot2 first though)!
theme_set(theme_bw(base_size=10))
theme_update(panel.background = element_rect(fill = "transparent",colour = NA),
             plot.background = element_rect(fill = "transparent",colour = NA))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

@@

```{r obs, dependson="sdp-pars"}
obs <- sim_obs(Xo, z_g, f, p, Tobs=Tobs, nz=1, 
                harvest = sort(rep(seq(0, .5, length=7), 5)), seed = seed_i)
```

@@

Once the posterior Gaussian process (GP) has been estimated [e.g. see
@Munch2005], it is necessary to adapt it in place of the parametric
equation for the stochastic dynamic programming (SDP) solution [see
@Mangel1988 for a detailed description of parametric SDP methods] to the
optimal policy. The essense of the idea is straight forward -- we will use
the estimated GP in place of the parametric growth function to determine
the stochastic transition matrix on which the SDP calculations are based.
The SDP is solved in a discritized state space -- both the continuously 
valued population densities $X$ and harvest quotas $h$ are first mapped to
a bounded, discrete grid.  (For simplicity we will consider a uniform grid,
though for either parametric or GP-based SDP it is often advantageous to 
use a non-uniform discretization such as a basis function representation,
e.g. see [@Deisenroth2009]).

The SDP approach then computes a transition matrix, $\mathbf{F}$.  We 
demonstrate that calculation is just as straight forward based on the GP
as it is in the classical context using the parametric model.  The 
${i,j}$ of the transition matrix $F$ entry gives the probability of transitioning into 
state $x_i$ given that the system is in state $x_j$ in the previous 
timestep.  To generate the transition
matrix based on the posterior GP, we need only the expected values
at each grid point and the corresponding variances (the diagonal of
the covariance matrix), as shown in Figure 1.  Given the mean of the 
GP posterior at each gridpoint as the vector $E$ and variance at that
point as vector $V$, the probability of transitioning from state $x_i$ to state $x_j$ is

$$\mathcal{N}\left(x_j | \mu = E_i, \sigma = \sqrt{V_i}\right)$$

where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$.  Strictly speaking, the transition probability should
be calculated by integrating the normal density over the bin of width
$\Delta$ centered at $x_j$.  For a sufficiently fine grid that $f(x_j)
\approx f(x_j + \Delta)$, it is sufficient to calculate the density at
$x_j$ and then row-normalize the transition matrix. The process
can then be repeated for each possible discrete value of our control 
variable, (harvest $h$).

**Pseudocode for the determining the transtion matrix from the GP**

@@

Using the discrete transition matrix we may write down the Bellman
recursion defining the the stochastic dynamic programming iteration:

@@

where $V(x_t)$ is the value of being at state $x$ at time $t$, $h$
is control (harvest level) chosen. Numerically, the maximization is
accomplished as follows. Consider the set of possible control values to
be the discrete values corresponding the the grid of
stock sizes.  Then for each $h_t$ there is a corresponding transition
matrix $\mathbf{F}_h$ determined as described above but with mean 
$\mu = x_j - h_t$. Let $\vec{V_t}$ be the vector whose $i$th element corresponds
to the value of having stock $x_i$ at time $t$.  Then let $\Pi_h$ be
the vector whose $i$th element indicates the profit from harvesting
at intensity $h_t$ given a population $x_i$ (e.g. $\max(x_i, h_t)$
since one cannot harvest more fish then the current population size).
Then the Bellman recursion can be given in matrix form as

@@

This completes the algorithm adapting the GP to the sequential decision-making 
problem through SDP, which we believe has not yet been demonstrated in the
literature.  We provide an R package implementation of this, along with
the Gaussian process inference, in the supplemental materials.

@@

Concerns over the potential for tipping points in ecological dynamics
[@Scheffer2001] highlight the dangers of uncertainty in ecological
management and pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  Because intervention is often too late 
after a tipping point has been crossed (but see @Hughes2013), management
is most often concerned with avoiding potentially catostrophic tipping
points before any data is available at or following a transition that
would more clearly reveal these regime shift dyanmics [e.g. @Bestelmeyer2012].

<!-- Explain the connection to alternative stable states -->

To illustrate the value of the non-parametric Bayesian approach to management,
we focus on example of a system containing such a tipping point whose dynamics
can still be described by a simple, one-dimensional parametric model.  
We will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".

@@

@@

<!-- things that we could hit upon in discussion:
Using data simulated from a specified model rather than empirical data
permits the comparison against the true underlying dynamics, setting 
a bar for the optimal performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).
-->

<span style="color:#007FFF;">***Should we include any emprical examples?***</span>

### Estimating parametric models

The maximum likelihood estimates (MLE) will assume that the process noise $Z_t$
is log-normal, as in the simulation itself.  In reality the noise structure might 
differ from this assumption, making this a best-case scenario for the MLE approach.  
The GP must also make an assumption about the noise structure -- in this case we 
assume additve noise, though multiplicative noise would be more accurate.  This
will illustrate the flexibility of the GP in fitting different noise structures
even without knowledge of the correct assumptions.

The GP estimate is Bayesian approach which requires specification of
prior distributions for the hyperparameters [see Appendix for details,
or @Rassmussen1996, @Munch2005 for background], as given in Table 1.
Then given the data, both the GP and the MLE parametric models can
be estimated as described above, as we implement in R in the code
provided in supplementary materials.

@@

Figure 1 shows the  inferred Gaussian Process compared to the true and the MLE parametric
models. The grey shading shows the uncertainty estimated by the GP.  Unlike the MLE estimates,
the GP uncertainty changes depending on where the data is available, allowing for a much 
greater level of uncertainty at low stock sizes.  While it would be natural (and straight forward)
to condition the GP on passing through the origin (0,0) (see appendix), the estimate shown
here is based only on the observed data.

<!-- Would it be useful to show the uncertainty in the other models? -->

<!-- with (light grey) and without (darker grey) measurement error -->

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance (grey shading)  is also shown.  The training data is also shown as black points. "}
tgp_dat <- 
    data.frame(  x = gp$XX[[1]], 
                 y = gp$ZZ.km, 
                 ymin = gp$ZZ.km - 2 * sqrt(gp$ZZ.ks2), 
                 ymax = gp$ZZ.km + 2 * sqrt(gp$ZZ.ks2),
                 ymin2 = gp$ZZ.mean - 2 * sqrt(gp$ZZ.vark), 
                 ymax2 = gp$ZZ.mean + 2 * sqrt(gp$ZZ.vark))
  true <- sapply(x_grid, f, 0, p)
  alt_mean <- sapply(x_grid, alt$f, 0, alt$p)
  est_mean <- sapply(x_grid, est$f, 0, est$p)
  models <- data.frame(x=x_grid, GP=tgp_dat$y, 
                       Parametric=est_mean, 
                       True=true, 
                       Structural=alt_mean)
  models <- melt(models, id="x")
  names(models) <- c("x", "method", "value")

ggplot(tgp_dat) + 
#  geom_ribbon(aes(x,y,ymin=ymin,ymax=ymax), fill="gray80") +    # measurement error + process error
  geom_ribbon(aes(x,y,ymin=ymin2,ymax=ymax2), fill="gray60") + # process error only
  geom_line(data=models, aes(x, value, col=method), alpha=0.8, lwd=1) + 
  geom_point(data=obs, aes(x,y)) + 
  xlab(expression(X[t])) + ylab(expression(X[t+1])) +
  scale_colour_manual(values=cbPalette) 
```

@@

parameter       interpretation
---------       -------------- 
$\sigma^2$      The process noise (from the kernel)
$\tau^2$        Variance around the mean
$\beta_0$       The mean is given by a linear model of slope $\beta$ 
$d$             The length-scale of the covariance function
$n$             The observation error
